---
title: "FH_MicroHab"
format: html
editor: visual
---

# Fort Huachuca Micro-habitat Model

### 0) Libraries

```{r}
# Working Directory
library(here)
# Spatial core
library(terra)        # rasters & vectors
library(sf)           # vectors / I/O
# Data
library(dplyr)
# Download & pre-process Sentinel-2 (10 m)
library(sen2r)        # easiest way to get cloud-masked S2 L2A
# Modeling
library(ranger)       # fast random forest
# Optional: texture features
# install.packages("glcm")
library(glcm)
# Stratified survey design (GRTS)
library(spsurvey)

```

### 1) Imagery Download

#### A) Sentinel2 Data

```{r}
# AOI as sf polygon (replace with your study boundary)
aoi <- st_read(here("data/FH_Boundary.shp")) %>% st_transform(4326)

out_dir <- "data/sentinel2"
dir.create(out_dir, showWarnings = FALSE)

s2prod <- c("BOA", "SCL")  # bottom-of-atmosphere reflectance + Scene Classification Layer

safe_list <- sen2r(
  gui = FALSE,
  extent = aoi,
  timewindow = as.Date(c("2025-06-01","2025-08-31")),
  step_atmcorr = "l2a",
  list_prods = s2prod,
  mask_type = "cloud_and_shadow",
  max_cloud_safe = 20,
  processing_order = "by_groups",
  path_out = out_dir,
  path_l2a = file.path(out_dir, "SAFE"),  # 👈 tell sen2r where to put the SAFE files
  overwrite = TRUE
)

library(terra)
naip <- rast("path/to/naip_1m.tif")
plotRGB(naip, r=1, g=2, b=3, stretch="lin")

```

#### B) NAIP Data 30cm

```{r}
library(sf)
library(terra)
library(rstac)
library(purrr)
library(curl)

# --- Inputs ---------------------------------------------------------------
aoi     <- st_read(here::here("data/FH_Boundary.shp")) |> st_transform(4326)
aoi_buf <- st_buffer(aoi, dist = 500)                      # include edge tiles
bb      <- st_bbox(aoi_buf)
outdir  <- here::here("data/naip_tiles_2023")
dir.create(outdir, recursive = TRUE, showWarnings = FALSE)

# --- 1) Query STAC, fetch ALL pages (unsigned for now) -------------------
pc <- stac("https://planetarycomputer.microsoft.com/api/stac/v1")
items <- pc |>
  stac_search(collections = "naip",
              bbox = as.numeric(bb),
              datetime = "2023-01-01/2023-12-31",
              limit = 200) |>
  get_request() |>
  items_fetch()

# --- 2) Find which features intersect the AOI (we'll still download full tiles)
it_sf <- items_as_sf(items) |> st_make_valid()
hit   <- st_intersects(it_sf, st_union(aoi), sparse = FALSE)[,1]
sel   <- which(hit)
cat("Tiles intersecting AOI:", length(sel), "\n")

# --- Helpers --------------------------------------------------------------
safe_name <- function(feature, fallback_idx) {
  id <- feature$id
  if (is.null(id) || is.na(id)) id <- paste0("tile_", fallback_idx)
  id <- gsub("[^A-Za-z0-9_\\-]+", "_", id)
  paste0(id, ".tif")
}

pick_asset_href <- function(feature) {
  # Prefer a TIFF asset with role "data"
  a <- feature$assets
  if (is.null(a) || !length(a)) return(NULL)
  nm <- names(a)
  meta <- lapply(nm, function(k) {
    list(key   = k,
         href  = a[[k]]$href %||% NA_character_,
         type  = a[[k]]$type %||% NA_character_,
         roles = paste(a[[k]]$roles %||% character(), collapse = ","))
  })
  meta <- do.call(rbind, lapply(meta, as.data.frame, stringsAsFactors = FALSE))

  is_tif  <- grepl("image/tiff", meta$type, ignore.case = TRUE) |
             grepl("\\.tif(f)?($|\\?)", meta$href, ignore.case = TRUE)
  has_data <- grepl("data", meta$roles, ignore.case = TRUE)

  idx <- which(is_tif & has_data)
  if (!length(idx)) idx <- which(is_tif)
  if (!length(idx)) return(NULL)
  meta$href[idx[1]]
}

download_one_signed <- function(items_obj, idx, outdir, min_ok = 5e6, tries = 3) {
  # Rewrap a single feature as an items object and sign it fresh
  one <- items_obj
  one$features <- list(items_obj$features[[idx]])
  one <- items_sign_planetary_computer(one)

  ftr  <- one$features[[1]]
  href <- pick_asset_href(ftr)
  if (is.null(href)) {
    warning(sprintf("No TIFF asset for feature %d; skipping.", idx))
    return(NULL)
  }
  fn <- file.path(outdir, safe_name(ftr, idx))

  # Skip if already good
  if (file.exists(fn) && file.info(fn)$size > min_ok) return(fn)

  for (t in seq_len(tries)) {
   ok <- tryCatch({
  h <- curl::new_handle(
    low_speed_limit = 10000,      # 10 KB/s
    low_speed_time  = 60          # for at least 1 min
  )
  curl_download(url = href, destfile = fn, mode = "wb", quiet = FALSE, handle = h)
  TRUE
}, error = function(e) FALSE)

    if (ok && file.exists(fn) && file.info(fn)$size > min_ok) return(fn)

    # Re-sign before retry in case SAS token expired
    one <- items_sign_planetary_computer(one)
    ftr  <- one$features[[1]]
    href <- pick_asset_href(ftr)
    Sys.sleep(1)
  }
  warning(sprintf("Failed or tiny file for %s", basename(fn)))
  NULL
}

# --- 3) Iterate with a simple progress bar -------------------------------
paths <- vector("list", length(sel))
pb <- txtProgressBar(min = 0, max = length(sel), style = 3)
for (j in seq_along(sel)) {
  i <- sel[j]
  paths[[j]] <- download_one_signed(items, i, outdir)
  setTxtProgressBar(pb, j)
}
close(pb)

tif_paths <- Filter(Negate(is.null), paths)
cat("\nSaved tiles:", length(tif_paths), "\n")
print(basename(unlist(tif_paths)))

# --- 4) (Optional) Mosaic later; clip to AOI when ready ------------------
# rlist <- lapply(tif_paths, \(p) tryCatch(rast(p), error = \(e) NULL)) |> purrr::compact()
# full_mosaic <- do.call(mosaic, rlist)
# writeRaster(full_mosaic, here::here("output/naip_2023_full_mosaic.tif"), overwrite = TRUE)
# naip_clip <- crop(full_mosaic, vect(aoi)) |> mask(vect(aoi))
# writeRaster(naip_clip, here::here("output/naip_2023_clip.tif"), overwrite = TRUE)

```

### 2) Build a predictor stack

#### A) Sentinel2 (10m)

Compute key indices and (optionally) texture; keep layers at the same resolution/projection.

```{r}
# Read the BOA mosaic (bands: B02,B03,B04,B08 at 10m; add B11,B12 at 20m if you resample)
boa <- rast(list.files(out_dir, pattern = "BOA_10m.*tif$", recursive=TRUE, full.names=TRUE)) 
names(boa) <- c("B02","B03","B04","B08")  # blue, green, red, NIR (check order in your files)

# Indices
ndvi <- (boa$B08 - boa$B04) / (boa$B08 + boa$B04)
ndwi <- (boa$B03 - boa$B08) / (boa$B03 + boa$B08)     # McFeeters-like
savi <- ((boa$B08 - boa$B04) / (boa$B08 + boa$B04 + 0.5)) * (1.5)

names(ndvi) <- "NDVI"; names(ndwi) <- "NDWI"; names(savi) <- "SAVI"

# Optional: simple texture on NDVI (window = 3x3 or 5x5; larger = smoother)
ndvi_tex <- glcm:::glcm_cpp(as.matrix(ndvi, wide=TRUE), 
                            window=c(5,5), statistics=c("mean","contrast","entropy"))
# Convert back to SpatRaster
ndvi_tex_r <- rast(ndvi); values(ndvi_tex_r) <- ndvi_tex[,"mean"]; names(ndvi_tex_r) <- "NDVI_tex"

# Predictor stack
preds <- c(boa, ndvi, ndwi, savi, ndvi_tex_r)
preds <- mask(preds, vect(aoi))  # clip to AOI

```

#### B) NAIP (30cm)

```{r}
library(terra)
library(sf)
library(dplyr)

#--- Inputs you already have ---------------------------------------------------
# AOI polygon (WGS84)
aoi <- st_read(here::here("data/FH_Boundary.shp")) |> st_transform(4326)

# Either: a single NAIP mosaic you wrote earlier...
# naip <- rast(here::here("output/naip_2023_FH_clip.tif"))

# ...or a list of full tiles you downloaded. If you have multiple, mosaic them:
naip_files <- list.files(here::here("data/naip_fulltiles_2023"), "\\.tif(f)?$", full.names = TRUE)
if (length(naip_files) == 1) {
  naip <- rast(naip_files)
} else {
  rlist <- lapply(naip_files, \(f) tryCatch(rast(f), error = \(e) NULL))
  rlist <- Filter(Negate(is.null), rlist)
  stopifnot(length(rlist) > 0)
  naip <- do.call(mosaic, rlist)
}

#--- 1) Standardize NAIP band order to R,G,B,NIR -------------------------------
# Planetary Computer / AWS "analytic" NAIP is typically RGBN (1..4)
# Some sources are NRGB. We'll handle both. If unsure, set 'layout' manually.

guess_layout <- function(x) {
  if (nlyr(x) != 4) stop("Expected 4-band NAIP (RGBN). Got ", nlyr(x), " bands.")
  # Heuristic: NIR tends to have higher mean than Blue. Compare band means.
  m <- sapply(1:4, \(i) global(x[[i]], mean, na.rm=TRUE)[[1]])
  nir_idx <- which.max(m)              # NIR usually has the highest mean
  if (nir_idx == 4) return("RGBN")     # common layout
  if (nir_idx == 1) return("NRGB")     # other common layout
  # Fallback to RGBN (most common in PC/AWS analytic)
  "RGBN"
}

layout <- guess_layout(naip)  # or set layout <- "RGBN" / "NRGB" explicitly

idx <- switch(layout,
  "RGBN" = list(R=1, G=2, B=3, NIR=4),
  "NRGB" = list(NIR=1, R=2, G=3, B=4),
  stop("Unknown layout: ", layout)
)

naip4 <- c(naip[[idx$R]], naip[[idx$G]], naip[[idx$B]], naip[[idx$NIR]])
names(naip4) <- c("R","G","B","NIR")

#--- 2) Indices (on 8-bit reflectance; coerced to float automatically) --------
ndvi <- (naip4$NIR - naip4$R) / (naip4$NIR + naip4$R)
ndwi <- (naip4$G   - naip4$NIR) / (naip4$G   + naip4$NIR)   # McFeeters
savi <- 1.5 * (naip4$NIR - naip4$R) / (naip4$NIR + naip4$R + 0.5)

names(ndvi) <- "NDVI"; names(ndwi) <- "NDWI"; names(savi) <- "SAVI"

#--- 3) Optional texture (fast & memory-safe) ----------------------------------
# GLCM at 30 cm on a big AOI is heavy. Use a proxy texture via focal SD on NDVI.
# If you *really* want GLCM, downscale first (see comment below).
w <- matrix(1, 5, 5)  # ~1.5 m window at 30 cm
ndvi_sd <- focal(ndvi, w = w, fun = sd, na.policy = "omit", na.rm = TRUE)
names(ndvi_sd) <- "NDVI_sd"

# If you prefer true GLCM, downsample to ~2–5 m first, then compute:
# ndvi_coarse <- aggregate(ndvi, fact = 6)                     # 0.3 m * 6 ≈ 1.8 m
# ndvi_glcm_mean <- glcm::glcm(raster::raster(ndvi_coarse), window = c(5,5), statistics = "mean")
# ndvi_tex_r <- rast(ndvi_glcm_mean); names(ndvi_tex_r) <- "NDVI_glcm_mean"

#--- 4) Predictor stack & clip to AOI ------------------------------------------
preds <- c(naip4, ndvi, ndwi, savi, ndvi_sd)
preds <- mask(crop(preds, vect(aoi)), vect(aoi))   # precise clip

# Quick sanity checks
preds
res(preds)        # should be ~0.3 0.3
nlyr(preds)       # R,G,B,NIR + NDVI, NDWI, SAVI, NDVI_sd
plotRGB(naip4, r=1, g=2, b=3, stretch="lin")
plot(ndvi)        # expect vegetation bright

```

### 3) Classification

#### Option A: Unsupervised (discover microhabitats, then name classes in field)

```{r}
set.seed(42)
# sample pixels to fit kmeans efficiently
samp <- spatSample(preds, size = 50000, method = "random", na.rm = TRUE, as.points = FALSE)
X <- na.omit(as.data.frame(samp, cells=FALSE))

k <- 6  # choose number of microhabitat classes you expect
km <- kmeans(scale(X), centers = k, nstart = 10)

# predict cluster for all pixels
# terra::predict requires a model interface; do a manual classify:
pred_mat <- as.data.frame(preds, na.rm=FALSE)
pred_scaled <- scale(pred_mat)
cl_idx <- rep(NA_integer_, nrow(pred_scaled))
ok <- complete.cases(pred_scaled)
cl_idx[ok] <- stats::predict(km, pred_scaled[ok, , drop=FALSE])  # in recent R, kmeans has predict method; otherwise use closest center
cl_r <- preds[[1]]; values(cl_r) <- cl_idx
names(cl_r) <- "class_id"

```

#### Option B: Supervised RF (if you already have training polygons)

-   Prepare an **sf** layer with polygons covering representative patches; add a `class` attribute (factor).

<!-- -->

-   Extract training data, fit RF, predict.

```{r}
train_sf <- st_read("data/microhab_train.gpkg") %>% st_transform(crs(preds))
train_r  <- terra::extract(preds, vect(train_sf), df = TRUE)
train_df <- train_r %>% 
  left_join(train_sf %>% st_drop_geometry() %>% mutate(ID=1:n()), by="ID") %>%
  filter(!is.na(class)) %>%
  mutate(class = factor(class))

rf <- ranger(class ~ ., data = select(train_df, -ID), 
             num.trees = 500, mtry = floor(sqrt(ncol(train_df)-2)), 
             importance = "impurity", probability = FALSE)

class_r <- predict(preds, rf, type = "response", na.rm=TRUE)
names(class_r) <- "class_id"

```

### 4) Smooth & polygonize (vector polygons per class)

```{r}
# Optional: small modal filter to de-speckle (3x3)
w <- matrix(1, 3, 3)
class_f <- focal(class_r, w=w, fun=modal, na.policy="omit", na.rm=TRUE)

# Dissolve contiguous pixels into polygons; drop tiny slivers
polys <- as.polygons(class_f, dissolve=TRUE, values=TRUE, na.rm=TRUE)
polys <- st_as_sf(polys) %>% st_make_valid()
polys <- polys %>%
  mutate(area_m2 = as.numeric(st_area(geometry))) %>%
  filter(area_m2 >= 1000)  # keep polygons ≥ 0.1 ha (adjust)

```

Add human-readable class names

```{r}
lut <- tibble(class_id = 1:6,
              class_name = c("Riparian dense","Mesquite savanna","Shrub-steppe",
                             "Open wash","Grass-dominated","Bare/rock"))
polys <- polys %>% left_join(lut, by="class_id")

```

### 5) Allocate **stratified** ground-truth points (area-proportional)

#### A) Quick stratified (proportional to area) with a min per class

```{r}
# area by class
tab <- polys %>%
  st_drop_geometry() %>%
  group_by(class_id) %>%
  summarise(area_m2 = sum(area_m2), .groups="drop") %>%
  mutate(area_prop = area_m2/sum(area_m2))

N_total <- 200        # total points you want to visit
min_per <- 10         # ensure at least this many per class

tab <- tab %>%
  mutate(n_alloc = pmax(min_per, round(N_total * area_prop))) %>%
  mutate(n_alloc = as.integer(round(n_alloc * N_total / sum(n_alloc))))  # normalize to N_total

# sample points within each class (uniform within polygons)
set.seed(123)
pts_list <- lapply(seq_len(nrow(tab)), function(i){
  cid <- tab$class_id[i]
  n_i <- tab$n_alloc[i]
  poly_i <- polys %>% filter(class_id == cid)
  # spatSample on polygons
  terra::spatSample(vect(poly_i), size=n_i, method="random", as.points=TRUE) %>%
    st_as_sf() %>%
    mutate(class_id = cid)
})
gt_pts <- do.call(rbind, pts_list) %>%
  left_join(lut, by="class_id")

```

#### B) GRTS (Generalized Random Tessellation Stratified) design with `spsurvey`

GRTS gives excellent spatial balance within each stratum.

```{r}
# Prepare a stratum field
polys_strata <- polys %>% mutate(stratum = as.factor(class_id))

# Sample-size list by stratum (from tab)
strata_n <- tab$n_alloc; names(strata_n) <- as.character(tab$class_id)

# spsurvey expects polygons (frame) and a named vector for stratum sample sizes
grts_out <- grts(
  design       = strata_n,     # named int vector
  DesignID     = "Microhabitat_GRTS",
  type.frame   = "area",
  src.frame    = "sf.object",
  sf.object    = polys_strata,
  stratumID    = "stratum",
  mdcaty       = NULL,
  shapefile    = FALSE
)

gt_pts_grts <- st_as_sf(grts_out$sites_legacy_areas) %>%
  rename(class_id = stratum) %>%
  mutate(class_id = as.integer(as.character(class_id))) %>%
  left_join(lut, by="class_id")

```

### 6) Export layers for field work

```{r}
st_write(polys,        "output/microhabitat_polys.gpkg", layer="microhabitat_polys", delete_dsn=TRUE)
writeRaster(class_f,   "output/microhabitat_class_30cm.tif", overwrite=TRUE)
st_write(gt_pts,       "output/gt_points_stratified.gpkg", layer="gt_points", delete_dsn=TRUE)
st_write(gt_pts_grts,  "output/gt_points_grts.gpkg",       layer="gt_points_grts", delete_dsn=TRUE)

```

### 7) Field labeling → accuracy assessment (post-survey)

Once you collect reference labels at `gt_points*`, bring the points back, join to the classified raster, and compute a confusion matrix, user’s/producer’s accuracy, and Kappa.

```{r}
# Extract predicted class at ground-truth points
gt <- st_read("output/gt_points_grts_with_field_labels.gpkg")  # add a 'truth' field in the field app
pred_at_pts <- terra::extract(class_f, vect(gt))
cmp <- gt %>% 
  st_drop_geometry() %>%
  bind_cols(pred_at_pts %>% select(class_id_pred = class_id)) %>%
  mutate(truth = factor(truth),
         pred  = factor(class_id_pred, levels = levels(truth)))  # align levels

# Confusion matrix
table(cmp$truth, cmp$pred)

# Overall & per-class accuracy
overall_acc <- mean(cmp$truth == cmp$pred, na.rm = TRUE)
by_class <- cmp %>% group_by(truth) %>% summarise(ua = mean(truth == pred))
overall_acc; by_class

```
